---
title: "Problem xy"
author: "Candidate number "
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document
# html_document
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,  comment = "#>")
```


You can start with an R-chunk where you load some R packages, for example as follows (replace with the R packages you need to solve the task). But replace the chunk option by `echo=FALSE`, because it is not something you need to print in the pdf you upload.

```{r, echo = FALSE}
library(knitr)
library(MASS)
library(keras)
library(caret)
library(pls)
library(glmnet)
library(gam)
library(gbm)
library(randomForest)
```

REPLACE

```{r}
id <- "1nLen1ckdnX4P9n8ShZeU7zbXpLc7qiwt" # google file ID
d.worm <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))
head(d.worm)
```


## a)

R code, results and answers to sub-question a)


## b) 
Below you have to complete the code and then replace `eval=FALSE` by `eval=TRUE` in the chunk options:
```{r, eval=FALSE, echo=TRUE}
ggplot(d.worm,aes(x= ... ,y=  ... ,colour= ...)) + geom_point() + theme_bw()
```

Note that the default figure width and height have been set globally as `fig.width=4, fig.height=3`, but if you would like to change that (e.g., due to space constraints), you can include a different width and height directly into the chunk options, again using `fig.width=..., fig.height=...`.

## c)
Here is an example with LaTeX code: $y_i = \beta_0 + \beta_1 x_i + \epsilon_i$, where $\beta_0=1$ and $\beta_1= 2$. In display mode you use for example 

$$y_i = \beta_0 + \beta_1 x_i + \epsilon_i \ , \epsilon_i \sim \mathcal{N}(0,\sigma^2) \ .$$
