---
title: "Module 7: Recommended Exercises"
author: "alexaoh"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    df_print: kable
subtitle: Statistical Learning V2021
---

```{r setup, include=FALSE}
showsolA<-TRUE
showsolB<-TRUE
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=68),tidy=TRUE)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize", comment = "#>")
library(ggplot2)
library(gam)
```

## Problem 1

```{r}
library(ISLR)
# extract only the two variables from Auto
ds = Auto[c("horsepower","mpg")]
n = nrow(ds)
#which degrees we will look at
deg = 1:4
set.seed(1)
#training ids for training set
tr = sample.int(n = n, size = n/2)
# plot of training data
plot(ds[tr,], col = "darkgrey", main = "Polynomial regression")

colors <- rainbow(n = length(deg))

MSE = sapply(deg, function(d){
  fit <- lm(mpg~poly(horsepower, d), data = ds[tr, ])
  lines(sort(ds[tr, 1]), fit$fit[order(ds[tr, 1])], col = colors[d])
  
  return (mean((predict(fit, ds[-tr, ]) - ds[-tr, 2])^2))
})
legend("topright", legend=paste("d = ", deg), col = colors, lty = 1)

# Plot MSE
plot(MSE, type = "o", xlab= "Pol. Deg.", main = "Test Error (MSE)", pch = 16)
```

## Problem 2

```{r}
attach(Auto)
fit2 <- lm(mpg~factor(origin))
dframe <- data.frame(origin = as.factor(sort(unique(origin))))
pred <- predict(fit2, dframe, se = T)

# Data frame including CI (z_alpha/2 = 1.96).
dat <-  data.frame(origin = dframe, mpg = pred$fit, lwr = pred$fit - 1.96 * pred$se.fit, 
    upr = pred$fit + 1.96 * pred$se.fit)

# Plot the fitted/predicted values and CI
ggplot(dat, aes(x = origin, y = mpg)) + geom_point() + geom_segment(aes(x = origin, 
    y = lwr, xend = origin, yend = upr)) + scale_x_discrete(labels = c(`1` = "1.American", 
    `2` = "2.European", `3` = "3.Japanese"))
```

## Problem 3

Now, let us look at the `Wage` data set. The section on Additive Models [(slides 28-34 in the pdf)](https://github.com/stefaniemuff/statlearning/blob/master/7BeyondLinear/7slides.pdf) explains how we can create an AM by adding components together. One component we saw is a natural spline in `year` with one knot. Derive the expression for the design matrix $\mathbf X_2$ from the natural spline basis 

$$
b_1(x_i) = x_i, \quad b_{k+2}(x_i) = d_k(x_i)-d_K(x_i),\; k = 0, \ldots, K - 1,\\
$$
$$
d_k(x_i) = \frac{(x_i-c_k)^3_+-(x_i-c_{K+1})^3_+}{c_{K+1}-c_k}.
$$


From the slides that are referenced to above, we know that the design matrix $\mathbf X_2$ is 

$$
\mathbf X_2 = \begin{pmatrix}
x_{12} & \left[\frac16(x_{12}-2003)^3 - \frac13(x_{12} - 2006)^3_+\right]\\
x_{22} & \left[\frac16(x_{22}-2003)^3 - \frac13(x_{22} - 2006)^3_+\right]\\
\vdots & \vdots\\
x_{n2} & \left[\frac16(x_{n2}-2003)^3 - \frac13(x_{n2} - 2006)^3_+\right]\\
\end{pmatrix}, 
$$

when having a knot at $c_1 = 2006$ and boundary knots at $c_0 = 2003$ and $c_2 = 2009$. The reason behind this matrix is given in the following. 

Since we are using only one knot, we set $K = 1$. Moreover, the first column of the matrix is always given by the functions $b_1(x_i) = x_i$. Since $K = 1$, $k = 0$ is the only value that $k$ takes. Hence, 

$$
\begin{split}
b_2(x_i) &= d_0(x_i) - d_1(x_i) = \frac{(x_i-c_0)^3_+-(x_i-c_2)^3_+}{c_2-c_0} - \frac{(x_i-c_1)^3_+-(x_i-c_2)^3_+}{c_2-c_1} \\
&= \frac{(x_i-2003)^3_+-(x_i-2009)^3_+}{6} - \frac{(x_i-2006)^3_+-(x_i-2009)^3_+}{3} \\
&= \frac16(x_i-2003)^3_+ - \frac13(x_i-2006)^3_+ + \frac16(x_i-2009)^3_+.
\end{split}
$$

Furthermore, since $2003 \leq x_i \leq 2009$, $(x_i-2009)^3_+ = 0$. Hence, $b_2(x_i) =  \frac16(x_i-2003)^3_+ - \frac13(x_i-2006)^3_+$. Finally, the design matrix is constructed by setting 

$$
\mathbf X_2 = \begin{pmatrix}
b_1(x_1) & b_2(x_1)\\
b_1(x_2) & b_2(x_2)\\
\vdots & \vdots\\
b_1(x_n) & b_2(x_n)\\
\end{pmatrix}.
$$

## Problem 4

Continuation of Problem 3. Write code that produces $\mathbf X$. 

```{r}
attach(Wage)
#X_1
mybs = function(x,knots){
  cbind(x,x^2,x^3,sapply(knots,function(y) pmax(0,x-y)^3))
}
d = function(c, cK, x) (pmax(0,x-c)^3-pmax(0,x-cK)^3)/(cK-c)
#X_2
myns = function(x,knots){
  kn = c(min(x), knots, max(x))
  K = length(kn)
  sub = d(kn[K-1],kn[K],x)
  cbind(x,sapply(kn[1:(K-2)],d,kn[K],x)-sub)
}
#X_3
myfactor = function(x) model.matrix(~x)[,-1]

# Define the X-matrix below. 

knots.age <- c(40,60)
knot.year <- 2006

X <- cbind(1, mybs(age, knots.age), myns(year, knot.year), myfactor(education))

# fitted model with our X
myhat = lm(wage ~ X - 1)$fit
# fitted model with gam
yhat = gam(wage ~ bs(age, knots = c(40,60)) + ns(year, knots = 2006) + education)$fit
# are they equal?
all.equal(myhat,yhat)
# Yes, they are equal!
```


## Problem 5

```{r}
Auto$origin <- as.factor(Auto$origin)
gamobject <- gam(mpg ~ bs(displacement, knots = c(290)) + poly(horsepower, 2) + weight + s(acceleration, df = 3) + origin, data = Auto)
summary(gamobject)

par(mfrow=c(2,3))
plot(gamobject, se = TRUE, col = "blue")
```
